---
title: "Chapter 3: Attribute data operations"
author: "Zexuan Zhao"
date: "14 October 2018"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

```{r}
library(sf)
library(raster)
library(tidyverse)
library(spData)
```

# 3.1 Introduction

Attribute data is non-spatial information associated with geographic (geometry) data. 

# 3.2 Vector attribute manipulation

sf objects have one column per attribute variable (such as ‘name’) and one row per observation, or feature (e.g. per bus station). 

sf provides methods to allow sf objects to behave like regular data frame.
```{r}
methods(class = "sf")
```

Create sf object from data.frame object by adding a geometry column
```{r}
st_sf(data.frame(n = world$name_long), g = world$geom)
```

Convert sf object to data.frame using st_set_geometry

```{r}
world_df = st_set_geometry(world, NULL)
class(world_df)
```

### Pitfall 1: name clashes  
x tidyr::extract() masks raster::extract()
x dplyr::filter()  masks stats::filter()
x dplyr::lag()     masks stats::lag()
x dplyr::select()  masks raster::select()

### Pitfall 2: tidyverse and sp don’t play

* sp precedes sf
* Together with the rgdal and rgeos package, it creates a powerful tool to works with spatial data
* Many spatial R packages still depends on the sp package, therefore it is important to know how to convert sp to and from sf objects

```{r}
library(spData)
world_sp = as(world, "Spatial")
world_sf = st_as_sf(world_sp)
```  
* The structures in the sp packages are more complicated - str(world_sf) vs str(world_sp)
* sp doesn’t play well with the tidyverse:
```{r}
world_sp %>% 
  filter(name_long == "England")
```

### Pitfall 3: multipolygon objects

This pitfall is not specific to the tidyverse but is worth being aware of. Let’s create a buffer around London of 500 km:
```{r message = FALSE}
lnd_buff <-  lnd[1, ] %>% 
  st_transform(crs = 27700) %>%  # world CRS to uk CRS
  st_buffer(500000) %>% #create buffer
  st_transform(crs = 4326) # uk CRS to world CRS
near_lnd <-  world[lnd_buff, ] # find countries in world that intersect with the buffer around UK
b <- st_bbox(near_lnd)
world %>% 
  st_geometry() %>% 
  plot(xlim = b[c(1,3)], ylim = b[c(2,4)])

near_lnd %>% # plot it
  select(name_long) %>% 
  plot(add = TRUE)
```  
What is going with the country miles away?
The issue is that some objects have multiple geometries:

```{r}
st_geometry_type(near_lnd)
```

Find countries that have more than one type of geometries
```{r}
data.frame(near_lnd$name_long,
           sapply(near_lnd$geom, length))
```

We can resolve this issue by casting them:

```{r}
world_poly <-  world %>% 
  st_cast(to = "POLYGON")
near_lnd_new = world_poly[lnd_buff, ]

b_new <- st_bbox(near_lnd_new)
world_poly %>% 
  st_geometry() %>% 
  plot(xlim = b_new[c(1,3)], ylim = b_new[c(2,4)])

near_lnd_new %>% # plot it
  select(name_long) %>% 
  plot(add = TRUE)
```

### Pitfall 4: spatial subsetting

Base R subsetting:  
```{r}
near_lnd <-  world_poly[lnd_buff, ]
```

Tidy spatial subsetting:  
It’s verbose (you need sparse = FALSE in the spatial predicate function).
* sparse = FALSE returns a dense matrix with TRUE/FALSE in each cell
* sparse = TRUE returns a sparse matrix with each row specifying which columns are denoted as true in the dense matrix
```{r message = FALSE}
near_lnd_tidy <- world_poly %>% 
  filter(st_intersects(., lnd_buff, sparse = FALSE))
```

### Pitfall 5: row names

Tidy spatial subsetting boshes the row names!!!  
```{r}
row.names(near_lnd_tidy)
row.names(near_lnd)
```

### Pitfall 6: attribute alteration
```{r}
sf1 <- world %>% 
  filter(name_long == "United Kingdom")
sf2 <- world[world$name_long == "United Kingdom", ]
identical(sf1, sf2)
```

Even after resetting row names, their still different.  
```{r}
row.names(sf1) <- row.names(sf2)
identical(sf1, sf2)
```

Advanced challenge: how to make u1 and u2 identical?  

```{r}
```

### Pitfall 7: Duplicate column names
```{r}

world_tidy <- st_as_sf(as_tibble(world))
st_intersection(world_tidy, lnd_buff) # fails
st_intersection(world, lnd_buff) # works
```

### Pitfall 8: binding rows

```{r}
rbind(near_lnd, near_lnd) # works
bind_rows(near_lnd, near_lnd) # fails
```

But you can do this:  
```{r}
near_lnd_data = st_set_geometry(near_lnd, NULL)
d = bind_rows(near_lnd_data, near_lnd_data)
d_sf = st_sf(d, geometry = c(near_lnd$geom, near_lnd$geom))
plot(d_sf)
```

## 3.2.1 Vector attribute subsetting  
Base R subsetting functions include [, subset() and $. dplyr subsetting functions include select(), filter(), and pull(). Both sets of functions preserve the spatial components of attribute data in sf objects.

Using []:  
```{r}
world[1:6, ] # subset rows by position
world[, 1:3] # subset columns by position
world[, c("name_long", "lifeExp")] # subset columns by name
```

```{r}
sel_area <-  world$area_km2 < 10000
summary(sel_area) # a logical vector
small_countries <-  world[sel_area, ] # filter by corresponding boolean
```

Using basic R function subset():  

```{r}
small_countries <-  subset(world, area_km2 < 10000)
```

Using dplyr::select():  

```{r}
# individually
world1 <-  dplyr::select(world, name_long, pop)
names(world1)
# all columns between name_long and pop (inclusive)
world2 <-  dplyr::select(world, name_long:pop)
names(world2)
# all columns except subregion and area_km2 (inclusive)
world3 <-  dplyr::select(world, -subregion, -area_km2)
names(world3)
# rename column in select()
world4 <-  dplyr::select(world, name_long, population = pop)
names(world4)
```

All dplyr functions including select() always return a data frame-like object. To extract a single vector, one has to explicitly use the pull()/$ command. 

```{r}
pull(world, pop)
world$pop
```

The subsetting operator [] in base R, by contrast, tries to return objects in the lowest possible dimension. This means selecting a single column returns a vector in base R. To turn off this behavior, set the drop argument to FALSE.

```{r}
# create throw-away data frame
d <-  data.frame(pop = 1:10, area = 1:10)
# return data frame object when selecting a single column
d[, "pop", drop = FALSE] # equivalent to d["pop"]
select(d, pop)
# return a vector when selecting a single column
d[, "pop"]
pull(d, pop)
```

Using slice() to select rows by index:  

```{r}
slice(world, 3:5)
```

Using filter() to select rows by matching given criteria

```{r}
world6 <-  filter(world, lifeExp > 82)
```

## 3.2.2 Vector attribute aggregation  

Aggregation operations summarize datasets by a ‘grouping variable’, typically an attribute column.  

Using base R function aggregate():  
```{r}
world_agg1 <-  aggregate(pop ~ continent, FUN = sum, data = world, na.rm = TRUE)
class(world_agg1)
```   
The result is a non-spatial data frame with six rows, one per continent, and two columns reporting the name and population of each continent.  

Using sf package function aggregate() when specifying by = :  
```{r}
world_agg2 <-  aggregate(world["pop"], by = list(world$continent),
                       FUN = sum, na.rm = TRUE)
class(world_agg2)
```  
This time the result is an sf object.

Using dplyr package function summarize():  
```{r}
world_agg3 <-  world %>%
  group_by(continent) %>%
  summarize(pop = sum(pop, na.rm = TRUE))
```

Finding the world's 3 most populous continents:  
```{r}
world %>% 
  dplyr::select(pop, continent) %>% 
  group_by(continent) %>% 
  summarize(pop = sum(pop, na.rm = TRUE), n_countries = n()) %>% 
  top_n(n = 3, wt = pop) %>%
  st_set_geometry(value = NULL) 
```

## 3.2.3 Vector attribute joining

dplyr join functions work the same on data frames and sf objects, the only important difference being the geometry list column. The result of data joins can be either an sf or data.frame object. The most common type of attribute join on spatial data takes an sf object as the first argument and adds columns to it from a data.frame specified as the second argument.

If you join a tibble with an sf object, it will return a tibble, with no geometry:  
```{r}
coffee_world = left_join(coffee_data, world)
class(coffee_world)
```

Normally we join an sf object with a tibble to add more attributes. It will return an sf object:  
```{r}
world_coffee <-  left_join(world, coffee_data)
plot(world_coffee["coffee_production_2017"])
```

Notice that there is a difference in country names:  

```{r}
setdiff(coffee_data$name_long, world$name_long)
```

The result shows that Others accounts for one row not present in the world dataset and that the name of the Democratic Republic of the Congo accounts for the other: it has been abbreviated, causing the join to miss it. 

The following command uses string matching (regex) to confirm what Congo, Dem. Rep. of should be:  
```{r}
str_subset(world$name_long, "Dem*.+Congo")
```

To fix this issue we will create a new version of coffee_data and update the name. inner_join()ing the updated data frame returns a result with all 46 coffee producing nations:  
```{r}
coffee_data$name_long[grepl("Congo,", coffee_data$name_long)] <- # search the cell to change
  str_subset(world$name_long, "Dem*.+Congo")
world_coffee_match <-  inner_join(world, coffee_data)
nrow(world_coffee_match)
```

## 3.2.4 Creating attributes and removing spatial information

Using base R:
```{r}
world_new = world # do not overwrite our original data
world_new$pop_dens = world_new$pop / world_new$area_km2
```

Using dplyr function mutate() or transmutate():  
```{r}
world %>% 
  mutate(pop_dens = pop / area_km2)
world %>% 
  transmute(pop_dens = pop / area_km2) # skips all other existing columns 
```

unite() pastes together existing columns. Additionally, we can define a separator (here: a colon :) which defines how the values of the input columns should be joined, and if the original columns should be removed (here: TRUE):

```{r}
world_unite = world %>%
  unite("con_reg", continent:region_un, sep = ":", remove = TRUE)
```

The separate() function does the opposite of unite(): it splits one column into multiple columns using either a regular expression or character positions.  
```{r}
world_separate = world_unite %>% 
  separate(con_reg, c("continent", "region_un"), sep = ":")
```

Using rename() to rename columns:  
```{r}
world %>% 
  rename(name = name_long)
```

Using set_names() to change all column names at once, and requires a character vector with a name matching each column.   
```{r}
new_names <-  c("i", "n", "c", "r", "s", "t", "a", "p", "l", "gP", "geom")
world %>% 
  set_names(new_names)
```

# 3.3 Manipulating raster objects

In contrast to the vector data model underlying simple features (which represents points, lines and polygons as discrete entities in space), raster data represent continuous surfaces.  
Creating new raster:  
```{r}
elev <-  raster(nrows = 6, ncols = 6, res = 0.5,
              xmn = -1.5, xmx = 1.5, ymn = -1.5, ymx = 1.5,
              vals = 1:36)
```

The result is a raster object with 6 rows and 6 columns (specified by the nrow and ncol arguments), and a minimum and maximum spatial extent in x and y direction (xmn, xmx, ymn, ymax). The vals argument sets the values that each cell contains: numeric data ranging from 1 to 36 in this case. Raster objects can also contain categorical values of class logical or factor variables in R.  

```{r}
grain_order <-  c("clay", "silt", "sand") # set
grain_char <-  sample(grain_order, 36, replace = TRUE) # sample
grain_fact <-  factor(grain_char, levels = grain_order) # convert to factor as raster object can not contain characters
grain <-  raster(nrows = 6, ncols = 6, res = 0.5, 
               xmn = -1.5, xmx = 1.5, ymn = -1.5, ymx = 1.5,
               vals = grain_fact)
plot(grain)
```

As shown above, raster objects represent categorical variables, e.g. factor, as integers, so grain[1, 1] returns a number that represents a unique identifier, rather than “clay”, “silt” or “sand”. The raster object stores the corresponding look-up table or “Raster Attribute Table” (RAT) as a data frame in a new slot named attributes.  
To see RAT:  
```{r}
ratify(grain)
levels(grain)
```

To pull out the value of each identifier:  
```{r}
factorValues(grain, grain[c(1, 11, 35)])
```

To plot with categorical legend:  
```{r}
library(tmap)

colfunc <-  colorRampPalette(c("lightyellow", "rosybrown"))
colfunc2 <-  c("clay" = "brown", "silt" = "sandybrown", "sand" = "rosybrown")

p1 <-  tm_shape(elev) + 
  tm_raster(legend.show = TRUE, palette = colfunc(36), style = "cont", title = "") +
  tm_layout(outer.margins = rep(0.01, 4), 
            inner.margins = rep(0, 4)) +
  tm_legend(bg.color = "white")

p2 <-  tm_shape(as(grain, "SpatialGridDataFrame")) + 
  tm_raster(legend.show = TRUE, palette = colfunc2, title = "") +
  tm_layout(outer.margins = rep(0.01, 4), 
            inner.margins = rep(0, 4)) +
  tm_legend(bg.color = "white")

tmap_arrange(p1, p2, nrow = 1)
```

## 3.3.1 Raster subsetting
Raster subsetting is done with the base R operator [, which accepts a variety of inputs:  
*Row-column indexing
Cell IDs
Coordinates
Another raster object*

```{r}
# row 1, column 1
elev[1, 1]
# cell ID 4
elev[4]
# extract all values or complete rows
values(elev) 
getValues(elev)
getValues(elev, 2) # extract values in the second row
getValues(elev, 2, 2) # extract 2 rows of values starting from the second row
```

For multi-layered raster objects stack or brick, if you extract values from cells, it will return the cell value(s) for each layer, and also you can extract values from layers.

```{r}
r_stack = stack(elev, grain)
names(r_stack) = c("elev", "grain")
# extract 2 values from 2 layers in the first cell
r_stack[1]
# extract values from layer "elev"
raster::subset(r_stack, "elev") 
r_stack[["elev"]]
r_stack$elev
```

## 3.3.2 Summarizing raster objects

If you provide the summary() and cellStats() functions with a raster stack or brick object, they will summarize each layer separately, as can be illustrated by running:  

```{r}
summary(brick(elev, grain))
cellStats(brick(elev, grain), sd)
cellStats(brick(elev, grain), mean)
```

Raster value statistics can be visualized in a variety of ways. Specific functions such as boxplot(), density(), hist() and pairs() work also with raster objects:  
```{r}
boxplot(elev)
density(elev)
hist(elev)
```

# 3.4 Exercise  
```{r}
library(spData)
data(us_states) # sf object, 
data(us_states_df) # data.frame
```

1.Create a new object called us_states_name that contains only the NAME column from the us_states object. What is the class of the new object and what makes it geographic?  

```{r}
us_state_name <- us_states %>% 
  select(NAME)
class(us_state_name)
us_state_name
```

2.Select columns from the us_states object which contain population data. Obtain the same result using a different command (bonus: try to find three ways of obtaining the same result). Hint: try to use helper functions, such as contains or starts_with from dplyr  

```{r}
us_state_pop <- us_states %>% 
  select(starts_with("total"))
```

3.Find all states with the following characteristics (bonus find and plot them):
*Belong to the Midwest region.
Belong to the West region, have an area below 250,000 km2 and in 2015 a population greater than 5,000,000 residents (hint: you may need to use the function units::set_units() or as.numeric()).
Belong to the South region, had an area larger than 150,000 km2 or a total population in 2015 larger than 7,000,000 residents.*

```{r}
us_states %>% 
  filter(REGION == "Midwest")

us_states %>% 
  filter(REGION == "West") %>% 
  filter(unclass(AREA) < 250000) %>% 
  filter(total_pop_15 > 5000000)

us_states %>%
  filter(REGION == "South") %>% 
  filter(unclass(AREA) > 150000 | total_pop_15 > 7000000)
```

4.What was the total population in 2015 in the us_states dataset? What was the minimum and maximum total population in 2015?

```{r}
us_states %>% 
  summarize(total_2015 = sum(total_pop_15), 
            min_2015 = min(total_pop_15),
            max_2015 = max(total_pop_15)) %>% 
  st_set_geometry(NULL)
```

5.How many states are there in each region?

```{r}
us_states %>% 
  count(REGION) %>% 
  st_set_geometry(NULL)
```

6.What was the minimum and maximum total population in 2015 in each region? What was the total population in 2015 in each region?

```{r}
us_states %>% 
  group_by(REGION) %>% 
  summarize(total_2015 = sum(total_pop_15),
             min_2015 = min(total_pop_15),
             max_2015 = max(total_pop_15)) %>% 
  st_set_geometry(NULL)
```

7. Add variables from us_states_df to us_states, and create a new object called us_states_stats. What function did you use and why? Which variable is the key in both datasets? What is the class of the new object?

```{r}
us_states_stats <- us_states %>%  
  left_join(us_states_df, by = c("NAME" = "state"))
```

8.us_states_df has two more observations than us_states. How can you find them? (hint: try to use the dplyr::anti_join() function)

```{r}
anti_join(us_states_df, us_states, by = c("state" = "NAME"))
```

9.What was the population density in 2015 in each state? What was the population density in 2010 in each state?  

```{r}
us_state_density <- us_states %>% 
  mutate(density_2010 = total_pop_10 / unclass(AREA)) %>% 
  mutate(density_2015 = total_pop_15 / unclass(AREA))
```

10. How much has population density changed between 2010 and 2015 in each state? Calculate the change in percentages and map them.

```{r}
us_state_density %>% 
  mutate(gap = (density_2015 - density_2010) / density_2010) %>% 
  ggplot(aes(x = NAME, y = gap)) +
    geom_col() +
  coord_flip()
```

11.Change the columns names in us_states to lowercase.  

```{r}
names(us_states) <- tolower(names(us_states))
```

12.Using us_states and us_states_df create a new object called us_states_sel. The new object should have only two variables - median_income_15 and geometry. Change the name of the median_income_15 column to Income.  

```{r}
us_states_sel <- us_states %>%
  left_join(us_states_df, by = c("name" = "state")) %>% 
  select(median_income_15) %>% 
  rename(Income = median_income_15)
```

13.Calculate the change in median income between 2010 and 2015 for each state. Bonus: what was the minimum, average and maximum median income in 2015 for each region? What is the region with the largest increase of the median income?

```{r}
us_states %>%
  left_join(us_states_df, by = c("name" = "state")) %>% 
  mutate(change = median_income_15 - median_income_10) %>% 
  ggplot(aes(x = name, y = change)) +
    geom_col()+
    coord_flip()
```

14.Create a raster from scratch with nine rows and columns and a resolution of 0.5 decimal degrees (WGS84). Fill it with random numbers. Extract the values of the four corner cells.

```{r}
my_raster <- raster(nrows = 9, ncols = 9, res = 0.5)
values(my_raster) <- rnorm(ncell(my_raster))
my_raster[1,1]
my_raster[1, ncol(my_raster)]
my_raster[nrow(my_raster), 1]
my_raster[nrow(my_raster), ncol(my_raster)]
```

15.What is the most common class of our example raster grain?  
```{r}
grain %>% 
cellStats(modal) %>%
  factorValues(grain, .)

```

16.Plot the histogram and the boxplot of the data(dem, package = "RQGIS") raster.  
```{r}
data(dem, package = "RQGIS")
hist(dem)
boxplot(dem)
```

