---
title: "Chapter 2"
author: "Zexuan Zhao"
date: "12 October 2018"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

Load data and package needed.

```{r message = FALSE}
library(sf)            # classes and functions for vector data
library(raster)        # classes and functions for raster data
library(spData)        # load geographic data
library(spDataLarge)   # load larger geographic data
library(tidyverse)     # load data manipulation
```

# 2.1 Introduction:
* The vector data model represents the world using points, lines and polygons. These have discrete, well-defined borders, meaning that vector datasets usually have a high level of precision.Vector data tends to dominate the social sciences because human settlements tend to have discrete borders.  
* The raster data model divides the surface up into cells of constant size. Raster datasets are the basis of background images used in web-mapping and have been a vital source of geographic data since the origins of aerial photography and satellite-based remote sensing devices. Rasters aggregate spatially specific features to a given resolution, meaning that they are consistent over space and scalable (many worldwide raster datasets are available).Raster often dominates in environmental sciences because of the reliance on remote sensing data.  

# 2.2 Vector Data
* The geographic vector model is based on points located within a coordinate reference system (CRS). Points can represent self-standing features (e.g. the location of a bus stop) or they can be linked together to form more complex geometries such as lines and polygons.  
* Coordinates consist of two numbers representing distance from an origin, usually in x and y dimensions.

## 2.2.1 An introduction to simple features  
* Simple Features is a hierarchical data model that represents a wide range of geometry types.  
* Simple feature objects in R are stored in a data frame, with geographic data occupying a special column, a ‘list-column’. This column is usually named ‘geom’ or ‘geometry’.

Now we're using world dataset from spData
```{r}
head(world)
plot(world)
```

Notice that $plot()$ will create multiple maps, one for each variable in the world datasets.

Being able to treat spatial objects as regular data frames with spatial powers has many advantages, especially if you are already used to working with data frames. You can use $summary()$ function to provide an overview of the variables in the $world$ object.

```{r}
summary(world$lifeExp)
print("-----------------------")
summary(world["lifeExp"])
```

* Notice that in the second line geometry is kept unless the user deliberately removes them.  
* The word MULTIPOLYGON in the summary output above refers to the geometry type of features (countries) in the world object. This representation is necessary for countries with islands such as Indonesia and Greece. Other geometry types are described in section

sf(Spatial DataFrame) objects are easy to subset. The code below shows its first two rows and three columns. The output shows two major differences compared with a regular data.frame: the inclusion of additional geographic data (geometry type, dimension, bbox and CRS information - epsg (SRID), proj4string), and the presence of final geometry column:  

```{r}
world_mini  <-  world[1:2, 1:3]
world_mini
```

## 2.2.2 Why simple features?
Why use the sf package when sp is already tried and tested
* Fast reading and writing of data.
* Enhanced plotting performance.
* sf objects can be treated as data frames in most operations.
* sf functions can be combined using %>% operator and works well with the tidyverse collection of R packages.
* sf function names are relatively consistent and intuitive (all begin with st_).

Conversion:  
```{r}
# sf -> sp
library(sp)
world_sp <-  as(world, Class = "Spatial")
# sp -> sf
world_sf <-  st_as_sf(world_sp, "sf")
```  

## 2.2.3 Basic map making
Basic maps are created in sf with plot(). By default this creates a multi-panel plot (like sp’s spplot()), one sub-plot for each variable of the object.A legend or ‘key’ with a continuous color is produced if the object to be plotted has a single variable.  
```{r}
# multi-panel plot, one plot for each variable
plot(world[3:6])
# plotting continuous variable
plot(world["pop"])
```

You can add layers by setting $ setting = TRUE $.
Note that $ reset = FALSE $
  reset: logical; if FALSE, keep the plot in a mode that allows adding further map elements; if TRUE restore original mode after plotting.
```{r}
world_asia <-  world %>% 
  filter(continent == "Asia") # filter by continent
asia <-  st_union(world_asia)
plot(world["pop"], reset = FALSE) # plot the background
plot(asia, add = TRUE, col = "red") # add a layer of ansia
```

## 2.2.4 base plot arguments

The graph below illustrates this flexibility by overlaying circles, whose diameters (set with cex =) represent country populations, on a map of the world. A basic version of the map can be created with the following commands.  
Note that :  
* $ of_largest = TRUE $ : logical; for st_centroid: if TRUE, return centroid of the largest (sub)polygon of a MULTIPOLYGON rather than of the whole MULTIPOLYGON.  
* $ cex = cex $ : symbol size  
```{r warning = FALSE}
plot(world["continent"], reset = FALSE) # plot the background
cex = sqrt(world$pop) / 10000 # create a tibble for rescaled population
world_cents = st_centroid(world, of_largest = TRUE) # create a sf object to store centroid location
plot(st_geometry(world_cents), add = TRUE, cex = cex) # add a layer of centroids with sizes corresponding to population sizes
```

sf’s plot method also has arguments specific to geographic data. $ expandBB $, for example, can be used plot an sf object in context: it takes a numeric vector of length four that expands the bounding box of the plot relative to zero in the following order: bottom, left, top, right. This is used to plot India in the context of its giant Asian neighbors, with an emphasis on China to the East, in the following code chunk.
Note that:  
* $ lwd $ to set line width

```{r}
india <- world %>% 
  filter(name_long == "India") # filter by country name
plot(st_geometry(india), expandBB = c(0, 0.2, 0.1, 1), col = "gray", lwd = 3) # plot the first layer of india (only geometry), expand the bbox of india from four directions, set the filling color and boarder line width.
plot(st_geometry(world_asia), add = TRUE) # add the background whose expansion is in line with the first layer
```  

## 2.2.5 Geometry types

Simple features in R can take on one of the 17 geometry types supported by the sf package. In this chapter we will focus on the seven most commonly used types: POINT, LINESTRING, POLYGON, MULTIPOINT, MULTILINESTRING, MULTIPOLYGON and GEOMETRYCOLLECTION.

* Point: POINT(5 2)  
* Linestring: a sequence of points with a straight line connecting the points, LINESTRING (1 5, 4 4, 4 1, 2 2, 3 2)  
* Polygon is a sequence of points that form a closed, non-intersecting ring, POLYGON ((1 5, 2 2, 4 1, 4 4, 1 5))  
** Note that polygon has the first and the last point of a polygon in the same coordinates.  
* Multiversion of sf above: 
** Multipoint - MULTIPOINT (5 2, 1 3, 3 4, 3 2)
** Multistring - MULTILINESTRING ((1 5, 4 4, 4 1, 2 2, 3 2), (1 2, 2 4))  
** Multipolygon - MULTIPOLYGON (((1 5, 2 2, 4 1, 4 4, 1 5), (0 2, 1 2, 1 3, 0 3, 0 2)))  
* Geometrycollection: GEOMETRYCOLLECTION (MULTIPOINT (5 2, 1 3, 3 4, 3 2), LINESTRING (1 5, 4 4, 4 1, 2 2, 3 2))  

## 2.2.6 Simple feature geometries (sfg)  
The sfg class represents the different simple feature geometry types in R: point, linestring, polygon (and their ‘multi’ equivalents, such as multipoints) or geometry collection. There are a set of functions to create simple feature geometry objects (sfg) from scratch if needed.
* A point - st_point()
* A linestring - st_linestring()
* A polygon - st_polygon()
* A multipoint - st_multipoint()
* A multilinestring - st_multilinestring()
* A multipolygon - st_multipolygon()
* A geometry collection - st_geometrycollection()

sfg objects can be created from three native data types:  
* A numeric vector - a single point e.g. c()  
```{r}
# note that we use a numeric vector for points
st_point(c(5, 2)) # XY point
st_point(c(5, 2, 3)) # XYZ point
st_point(c(5, 2, 1), dim = "XYM") # XYM point, needed to be specified
st_point(c(5, 2, 3, 1)) # XYZM point
```  
* A matrix - a set of points, where each row contains a point - a multipoint or linestring  
```{r}
# the rbind function simplifies the creation of matrices
## MULTIPOINT
multipoint_matrix <- rbind(c(5, 2), c(1, 3), c(3, 4), c(3, 2))
st_multipoint(multipoint_matrix)
## LINESTRING
linestring_matrix <- rbind(c(1, 5), c(4, 4), c(4, 1), c(2, 2), c(3, 2))
st_linestring(linestring_matrix)
```  
* A list - any other set, e.g. a multi-x or geometry collection
```{r}
## POLYGON
polygon_list  <-  list(rbind(c(1, 5), c(2, 2), c(4, 1), c(4, 4), c(1, 5)))
st_polygon(polygon_list)
## POLYGON with a hole
polygon_border  <-  rbind(c(1, 5), c(2, 2), c(4, 1), c(4, 4), c(1, 5))
polygon_hole <-  rbind(c(2, 4), c(3, 4), c(3, 3), c(2, 3), c(2, 4))
polygon_with_hole_list <- list(polygon_border, polygon_hole)
st_polygon(polygon_with_hole_list)

## MULTILINESTRING
multilinestring_list <- list(rbind(c(1, 5), c(4, 4), c(4, 1), c(2, 2), c(3, 2)), 
                            rbind(c(1, 2), c(2, 4)))
st_multilinestring((multilinestring_list))

## MULTIPOLYGON
multipolygon_list <- list(list(rbind(c(1, 5), c(2, 2), c(4, 1), c(4, 4), c(1, 5))),
                         list(rbind(c(0, 2), c(1, 2), c(1, 3), c(0, 3), c(0, 2))))
st_multipolygon(multipolygon_list)

## GEOMETRYCOLLECTION
gemetrycollection_list <- list(st_multipoint(multipoint_matrix),
                              st_linestring(linestring_matrix))
st_geometrycollection(gemetrycollection_list)
```

## 2.2.7 Simple feature columns (sfc)

One sfg object contains only a single simple feature geometry. A simple feature geometry column (sfc) is a list of sfg objects, which is additionally able to contain information about the coordinate reference system in use. For instance, to combine two simple features into one object with two features, we can use the $ st_sfc() $ function. This is important since sfc represents the geometry column in sf data frames:

```{r}
# sfc POINT
point1  <-  st_point(c(5, 2))
point2  <-  st_point(c(1, 3))
points_sfc <-  st_sfc(point1, point2)
points_sfc
```

In most cases, an sfc object contains objects of the same geometry type. Therefore, when we convert sfg objects of type polygon into a simple feature geometry column, we would also end up with an sfc object of type polygon, which can be verified with st_geometry_type(). Equally, a geometry column of multilinestrings would result in an sfc object of type multilinestring:

```{r}
# sfc POLYGON
polygon_list1 <-  list(rbind(c(1, 5), c(2, 2), c(4, 1), c(4, 4), c(1, 5)))
polygon1 <-  st_polygon(polygon_list1)
polygon_list2 <-  list(rbind(c(0, 2), c(1, 2), c(1, 3), c(0, 3), c(0, 2)))
polygon2 <-  st_polygon(polygon_list2)
polygon_sfc <-  st_sfc(polygon1, polygon2)
st_geometry_type(polygon_sfc)
```  
```{r}
# sfc MULTILINESTRING
multilinestring_list1  <-  list(rbind(c(1, 5), c(4, 4), c(4, 1), c(2, 2), c(3, 2)), 
                            rbind(c(1, 2), c(2, 4)))
multilinestring1  <-  st_multilinestring((multilinestring_list1))
multilinestring_list2  <-  list(rbind(c(2, 9), c(7, 9), c(5, 6), c(4, 7), c(2, 7)), 
                            rbind(c(1, 7), c(3, 8)))
multilinestring2  <-  st_multilinestring((multilinestring_list2))
multilinestring_sfc  <-  st_sfc(multilinestring1, multilinestring2)
st_geometry_type(multilinestring_sfc)
```  

It is also possible to create an sfc object from sfg objects with different geometry types:

```{r}
point_multilinestring_sfc = st_sfc(point1, multilinestring1)
st_geometry_type(point_multilinestring_sfc)
```

As mentioned before, sfc objects can additionally store information on the coordinate reference systems (CRS). To specify a certain CRS, we can use the epsg (SRID) or proj4string attributes of an sfc object. The default value of epsg (SRID) and proj4string is NA (Not Available), as can be verified with $ st_crs() $:

```{r}
st_crs(points_sfc)
```

All geometries in an sfc object must have the same CRS. We can add coordinate reference system as a crs argument of $ st_sfc() $. 

```{r}
# EPSG definition
points_sfc_wgs <-  st_sfc(point1, point2, crs = 4326)
st_crs(points_sfc_wgs)
# PROJ4STRING definition
st_sfc(point1, point2, crs = "+proj=longlat +datum=WGS84 +no_defs")
```

## 2.2.8 The sf class  
Objects of class sf represent such data by combining the attributes (data.frame) with the simple feature geometry column (sfc). They are created with $ st_sf() $ as illustrated below.  
```{r}
lnd_point <-  st_point(c(0.1, 51.5))                 # sfg object: the coordinates were used to create the simple feature geometry
lnd_geom <-  st_sfc(lnd_point, crs = 4326)           # sfc object: the geometry was converted into a simple feature geometry column with a CRS
lnd_attrib <-  data_frame(                           # data.frame object: attributes were stored in a data.frame
  name = "London",
  temperature = 25,
  date = as.Date("2017-06-21")
  )
lnd_sf <-  st_sf(lnd_attrib, geometry = lnd_geom)    # sf object: combine data.frame with the sfc object using st_sf()
class(lnd_sf)
```  
The result shows that sf objects actually have two classes, sf and data.frame. Simple features are simply data frames (square tables), but with spatial attributes (usually stored in a special geometry list-column in the data frame). This duality is central to the concept of simple features: most of the time a sf can be treated as and behaves like a data.frame. Simple features are, in essence, data frames with a spatial extension.

# 2.3 Raster Data

The geographic raster data model usually consists of a raster header and a matrix (with rows and columns) representing equally spaced cells (pixels).  
The raster header defines the coordinate reference system, the extent and the origin. The origin (or starting point) is frequently the coordinate of the lower-left corner of the matrix (the raster package, however, uses the upper left corner, by default). The header defines the extent via the number of columns, the number of rows and the cell size resolution. Hence, starting from the origin, we can easily access and modify each single cell by either using the ID of a cell or by explicitly specifying the rows and columns. This matrix representation avoids storing explicitly the coordinates for the four corner points (in fact it only stores one coordinate, namely the origin) of each cell corner as would be the case for rectangular vector polygons. This and map algebra makes raster processing much more efficient and faster than vector data processing. However, in contrast to vector data, the cell of one raster layer can only hold a single value. The value might be numeric or categorical.

## 2.3.1 An introduction to raster

The raster package supports raster objects in R. It provides an extensive set of functions to create, read, export, manipulate and process raster datasets. raster also lets you work on large raster datasets that are too large to fit into the main memory. In this case, raster provides the possibility to divide the raster into smaller chunks (rows or blocks), and processes these iteratively instead of loading the whole raster file into RAM.

```{r}
raster_filepath = system.file("raster/srtm.tif", package = "spDataLarge")
new_raster = raster(raster_filepath)
dim(new_raster) # return the number of rows, columns and layers
ncell(new_raster) # return the number of cells (pixels)
res(new_raster) # return the raster’s spatial resolution
extent(new_raster) # return the spatial extent
crs(new_raster) # return the coordinate reference system
inMemory(new_raster) # return whether the the raster data is stored in memory (the default) or on disk
```

Note: The current version of raster is not fully compatible with sf objects. We therefore suggest converting sf objects to an intermediate Spatial class (from the sp package) for compatibility between raster and vector data, for example crop(raster_obj, as(sf_object, "Spatial")

## 2.3.2 Basic map making  
```{r}
plot(new_raster)
```

## 2.3.3 Raster classes

The RasterLayer class represents the simplest form of a raster object, and consists of only one layer. The easiest way to create a raster object in R is to read-in a raster file from disk or from a server.

```{r}
class(new_raster)
```
The raster package supports numerous drivers with the help of rgdal. To find out which drivers are available on your system, run $ raster::writeFormats() $ and $ rgdal::gdalDrivers() $. Drivers are code that help Windows and MacOS recognize the physical components of your computer, like a printer, graphics card, or mouse.

Create your own raster using the $ raster() $ function. This is illustrated in the subsequent code chunk, which results in a new RasterLayer object. The resulting raster consists of 36 cells (6 columns and 6 rows specified by nrows and ncols) centered around the Prime Meridian and the Equator (see xmn, xmx, ymn and ymx parameters). The CRS is the default of raster objects: WGS84. This means the unit of the resolution is in degrees which we set to 0.5 (res). Values (vals) are assigned to each cell: 1 to cell 1, 2 to cell 2, and so on. Remember: raster() fills cells row-wise (unlike matrix()) starting at the upper left corner, meaning the top row contains the values 1 to 6, the second 7 to 12 etc.


```{r}
new_raster2 <-  raster(nrows = 6, ncols = 6, res = 0.5, 
                     xmn = -1.5, xmx = 1.5, ymn = -1.5, ymx = 1.5,
                     vals = 1:36)
plot(new_raster2)
```

Aside from RasterLayer, there are two additional classes: RasterBrick and RasterStack. Both can handle multiple layers, but differ regarding the number of supported file formats, type of internal representation and processing speed.

A RasterBrick consists of multiple layers, which typically correspond to a single multispectral satellite file or a single multilayer object in memory. The brick() function creates a RasterBrick object. Usually, you provide it with a filename to a multilayer raster file but might also use another raster object and other spatial objects (see ?brick for all supported formats).

```{r}
multi_raster_file <-  system.file("raster/landsat.tif", package = "spDataLarge")
r_brick  <-  brick(multi_raster_file)
class(r_brick)
nlayers(r_brick) # nlayers() retrieves the number of layers stored in a Raster* object:
```

A RasterStack is similar to a RasterBrick in the sense that it consists also of multiple layers. However, in contrast to RasterBrick, RasterStack allows you to connect several raster objects stored in different files or multiply objects in memory. More specifically, a RasterStack is a list of RasterLayer objects with the same extent and resolution. Hence, one way to create it is with the help of spatial objects already existing in R’s global environment. And again, one can simply specify a path to a file stored on disk.

```{r}
raster_on_disk <-  raster(r_brick, layer = 1)
raster_in_memory <-  raster(xmn = 301905, xmx = 335745,
                          ymn = 4111245, ymx = 4154085, 
                          res = 30)
values(raster_in_memory) <-  sample(seq_len(ncell(raster_in_memory))) # set the value of stack as randomly sampled numbers from vectors
crs(raster_in_memory) <-  crs(raster_on_disk) # set the crs of stack the same as raster
r_stack <-  stack(raster_in_memory, raster_on_disk) # connect raster objects
plot(r_stack)
```

Processing of a single mulitilayer file or object is the most effective with RasterBrick, while RasterStack allows calculations based on many files, many Raster* objects, or both.

# 2.4 Coordinate Reference System  
Coordinate Reference System (CRS) defines how the spatial elements of the data relate to the surface of the Earth (or other bodies).

# 2.4.1 Geographic coordinate systems  
Geographic coordinate systems identify any location on the Earth’s surface using two values — longitude and latitude. Longitude is location in the East-West direction in angular distance from the Prime Meridian plane. Latitude is angular distance North or South of the equatorial plane. Distance in geographic CRSs are therefore not measured in meters.

The surface of the Earth in geographic coordinate systems is represented by a spherical or ellipsoidal surface. Spherical models assume that the Earth is a perfect sphere of a given radius. Spherical models have the advantage of simplicity but are rarely used because they are inaccurate: the Earth is not a sphere! Ellipsoidal models are defined by two parameters: the equatorial radius and the polar radius. These are suitable because the Earth is compressed: the equatorial radius is around 11.5 km longer than the polar radius

Ellipsoids are part of a wider component of CRSs: the datum. This contains information on what ellipsoid to use (with the ellps parameter in the PROJ CRS library) and the precise relationship between the Cartesian coordinates and location on the Earth’s surface. These additional details are stored in the towgs84 argument of proj4string notation (see proj4.org/parameters.html for details). These allow local variations in Earth’s surface, e.g. due to large mountain ranges, to be accounted for in a local CRS. There are two types of datum — local and geocentric. In a local datum such as NAD83 the ellipsoidal surface is shifted to align with the surface at a particular location. In a geocentric datum such as WGS84 the center is the Earth’s center of gravity and the accuracy of projections is not optimized for a specific location. Available datum definitions can be seen by executing st_proj_info(type = "datum").  

## 2.4.2 Projected coordinate systems  
Projected CRSs are based on Cartesian coordinates on an implicitly flat surface. They have an origin, x and y axes, and a linear unit of measurement such as meters. All projected CRSs are based on a geographic CRS, described in the previous section, and rely on map projections to convert the three-dimensional surface of the Earth into Easting and Northing (x and y) values in a projected CRS.  

This transition cannot be done without adding some distortion. Therefore, some properties of the Earth’s surface are distorted in this process, such as area, direction, distance, and shape. A projected coordinate system can preserve only one or two of those properties. Projections are often named based on a property they preserve: equal-area preserves area, azimuthal preserve direction, equidistant preserve distance, and conformal preserve local shape.  

## 2.4.3 CRSs in R

An epsg code is usually shorter, and therefore easier to remember. The code also refers to only one, well-defined coordinate reference system. 
A proj4string definition allows you more flexibility when it comes to specifying different parameters such as the projection type, the datum and the ellipsoid.

Spatial R packages support a wide range of CRSs and they use the long-established PROJ library. Other than searching for EPSG codes online, another quick way to find out about available CRSs is via the rgdal::make_EPSG() function, which outputs a data frame of available projections.

```{r}
crs_data = rgdal::make_EPSG()
View(crs_data)
```

In sf the CRS of an object can be retrieved using $ st_crs() $.

```{r}
vector_filepath <-  system.file("vector/zion.gpkg", package = "spDataLarge")
new_vector <-  st_read(vector_filepath)
st_crs(new_raster)
```

In cases when a coordinate reference system (CRS) is missing or the wrong CRS is set, the $ st_set_crs() $ function can be used:

```{r}
new_vector <-  st_set_crs(new_vector, 26912) # set CRS
```

The projection() function can be used to access/set CRS information from a Raster* object:

```{r}
projection(new_raster) # get CRS
projection(new_raster) <-  "+proj=utm +zone=12 +ellps=GRS80 +towgs84=0,0,0,0,0,0,0 
                            +units=m +no_defs" # set CRS
```

# 2.5 Units

An important feature of CRSs is that they contain information about spatial units. Clearly it is vital to know whether a house’s measurements are in feet or meters, and the same applies to maps. It is good cartographic practice to add a scale bar onto maps to demonstrate the relationship between distances on the page or screen and distances on the ground. Likewise, it is important to formally specify the units in which the geometry data or pixels are measured to provide context, and ensure that subsequent calculations are done in context.  

A novel feature of geometry data in sf objects is that they have native support for units. This means that distance, area and other geometric calculations in sf return values that come with a units attribute, defined by the units package. This is advantageous because it prevents confusion caused by the fact that different CRSs use different units (most use meters, some use feet). Furthermore, it also provides information on dimensionality, as illustrated by the following calculation which reports the area of Luxembourg:

```{r}
luxembourg  <-  world %>% 
  filter(name_long == "Luxembourg")
area <- st_area(luxembourg)
area
```

The result is in units of square meters (m2), showing that the result represents two-dimensional space. This information, stored as an attribute.

```{r}
attributes(area)
```
Reset Unit:

```{r}
units::set_units(st_area(luxembourg), km^2)
```

Units are of equal importance in the case of raster data. However, so far sf is the only spatial package that supports units, meaning that people working on raster data should approach changes in the units of analysis (for example, converting pixel widths from imperial to decimal units) with care. The new_raster object (see above) uses a WGS84 projection with decimal degrees as units. Consequently, its resolution is also given in decimal degrees but you have to know it, since the res() function simply returns a numeric vector.

```{r}
res(new_raster)
```

If we used the UTM projection, the units would change.

```{r warning = FALSE}
repr <-  projectRaster(new_raster, crs = "+init=epsg:26912")
res(repr)
```

# 2.6 Exercises
1.Use summary() on the geometry column of the world data object. What does the output tell us about:
Its geometry type? Multi-polygon
The number of countries? 177
Its coordinate reference system (CRS)? +proj=longlat +datum=WGS84 +no_defs

```{r}
summary(st_geometry(world))
st_crs(world)
```

2.Run the code that ‘generated’ the map of the world in Figure at the end of 2.2.4. Find two similarities and two differences between the image on your computer and that in the book.
What does the cex argument do ? set the symbol size
Why was cex set to the sqrt(world$pop) / 10000? to rescale
Bonus: experiment with different ways to visualize the global population.


3.Use plot() to create maps of Nigeria in context (see section 2.2.4).
Adjust the lwd, col and expandBB arguments of plot().
Challenge: read the documentation of text() and annotate the map.
```{r warning = FALSE}
# data preparation
# data wrangling
nigeria <-  world %>% 
  filter(name_long == "Nigeria")
world_africa <- world %>% 
  filter(continent == "Africa")

# calculate the coordinate where to put text
ncentre = st_centroid(nigeria)
ncentre_num = st_coordinates(ncentre) #returns x,y coordinates of the geometry

# add first layer Africa
plot(st_geometry(world_africa), col = "white", lwd = 3, main = "Nigeria in context", border = "lightgrey")
# add second layer World
plot(st_geometry(world), lty = 3, add = TRUE, border = "grey")
# add third layer Nigeria
plot(st_geometry(nigeria), col = "yellow", add = TRUE, border = "darkgrey")
# specify x,y coordinates of text
text(x = ncentre_num[1], y = ncentre_num[2], labels = "Nigeria")
```


4.Create an empty RasterLayer object called my_raster with 10 columns and 10 rows. Assign random values between 0 and 10 to the new raster and plot it.

```{r}
my_raster <- raster(nrow = 10, ncol = 10)
values(my_raster) <- 1:ncell(my_raster)
plot(my_raster)
```

5.Read-in the raster/nlcd2011.tif file from the spDataLarge package. What kind of information can you get about the properties of this file?

```{r}
filepath <-  system.file("raster/nlcd2011.tif", package = "spDataLarge")
raster(filepath)
```

